---
title: "Design"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Design}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(vizid)
```

## Data Frames are Primary Objects

Many functions in this package are **transformers**: they take a data frame and return a modified data frame. The data frame acts as the single source of truth for both your **data** and the **state** of your visualization session.

In this context, *state* refers to information stored in attributes—such as:

- which variables should be used for which roles in the visualization (e.g., the `cases` variable should be the `series_variable` used in a time-series plot).
- `ggplot2` components that will be combined to produce a plot object.
- other metadata captured during previous steps that influences default plotting behaviour.

Because transformers carry this state forward, you use them in a pipeline that ultimately produces a data frame ready for plotting.

```{r, eval = FALSE}
vizid_object <- (data
  # transformers from other packages that you love
  |> non_vizid_functions()
  |> ...
  
  # transformers that modify data *and* carry forward state
  |> vizid_functions()     
  |> ...
)

data_frame_exploration_function(vizid_object)
plot(vizid_object)
```

One consequence of this design is that you can inspect the **data that will be plotted** using all the standard R tools for examining data frames (`head()`, `summary()`, dplyr verbs, etc.). This is intentional: the visible content of the data frame always reflects the data that `vizid` will plot. Other aspects of the session state—such as inferred variable roles, user-supplied choices, and metadata collected along the way—are stored in **attributes**, which ordinary data-frame tools ignore but which you can view with `attributes()` if needed.

If you want to insert your own transformation into a `vizid` pipeline, you can write a custom transformer and wrap it using the helper provided by `vizid`, ensuring that your function returns a modified data frame **and** preserves all relevant `vizid` attributes.

The final step in a `vizid` pipeline is another data frame—one that includes the information required to produce a plot. This aligns with base R behaviour: `plot(data)` works for any data frame. `vizid` simply ensures that, for historical infectious-disease data, the resulting plot is informative and context-aware while still respecting the explicit choices you make (e.g., selecting the correct series when multiple candidates exist).
